A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

@@

\newcommand\Data{\Types_{\mathrm{dt}}}
\newcommand\Codata{\Types_{\mathrm{cdt}}}
\newcommand\Nondata{\Types_{\mathrm{pln}}}

@@

Codatatypes, like datatypes, are freely generated by their constructors;
but for codatatypes, infinite constructor terms also denote values
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all finite and infinite ground constructor terms, and
only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{S}(\infty) \teq \infty$ hold without any additional
work, because both sides of the equality share the same infinite expansion
$\const{S}(\const{S}(\const{S}(\ldots)))$.

@@

In this paper, we introduce a unified decision procedure for both datatypes and
codatatypes. The procedure is described abstractly as a calculus
(Section~\ref{sec:the-calculus}). To our knowledge, this is the first decision
procedure for codatatypes. For the datatype case, it roughly follows the lines
of Barrett et al.\ \cite{barrett-et-al-2007}. Because datatypes and codatatypes
share many of the same properties, it is sensible to consider them together.

@@

%  * setting :
    * universal formulas
    * many-sorted logic
    * mutually (co)recursive types with constructors, selectors, and
      discriminators

  * codatatypes: from a theoretical and implementational point of view, like
      datatypes but:
    * infinite values (infinitely many nested constructors)
    * codatatypes are never empty (e.g. finite streams are rejected)

\[
      \keyw{codatatype}\; \,\ty{llist} \,=\, \const{LNil} \,\mid\, \const{LCons}(\ty{int},\, \ty{llist})
\]

    * no acyclicity (e.g. xs = lcons(0, xs) is satisfiable)
    * instead: bisimilarity check
    * (also: ``enumeration'', e.g. of streams)

  * motivation:
    * acyclicity and bisimilarity cannot be axiomatized, so they really belong in
      a decision procedure
    * other properties can be done more efficiently by a decision procedure

  * consider a natural fragment---datatypes as supported in modern SMT solvers
    and the SMT-LIB 2 standard, and codatatypes as their duals
    * mutual recursion, but no polymorphism, nested recursion

  * integrated with Nelson-Oppen

  * implemented in CVC4 as rewriting and a theory solver
    (\ref{sec:the-theory-solver})
    * precise strategy for applying inferences
    * about 2000 lines of code, among which 1600 are shared between datatypes
      and codatatypes

  * useful both for proving and for model finding
    * in particular, acyclicity and bisimilarity are necessary for some proofs
      (and cannot be axiomatized finitely)
    * and for model finding, without them we quickly get spurious models
    * explain how finite model finding works

  * benchmarking is often an issue -- esp. codatatypes
  * the decision procedure is evaluated on two sets of benchmarks
    (Section~\ref{sec:experimental-results}):
    * first set: hand-crafted examples
    * second set: generated from Isabelle theories using the Sledgehammer tool

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall

The proofs are part of a report.

\newpage

* SMT-Lib

@@

  * perhaps the most commonly used codatatype is that of lazy lists or sequences.
    using a syntax similar to Standard ML, Haskell, or SMT-LIB

$\mu$ notation

  * less briefly: codatatypes and why they are useful
    * in Agda, Coq, Matita -- recently also in Isabelle/HOL \cite{nipkow-et-al-2002}
    * but not in SMT-LIB 2 \cite{barrett-et-al-2010}

\paragraph{Related Work.}

    * point to Barrett et al. for SMT datatypes
      * about their own work, they say: "our focus is on generality and
        efficiency rather than immediacy of implementation"
      * (deal more directly with finite sorts than Barrett et al., Section 6.1)
    * SPASS-Pirate
    * additional ones since then (e.g. strings?)
    * anything about codatatypes?
      * proof assistants like Agda, Coq, etc. have them
      * Dafny, CoALP
      * also a lot of theoretical research, some of which is loosely connected,
        e.g. decision procedure for corecursive functions (Henning in Nijmegen)

    * ODDITY: Oppen 1980: single-constructor, recursive -- infinite values?

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{11}^1{:}\bigr]\; \tau_{11}^1, \,\ldots,\, \bigl[\const{s}_{11}^{n_{11}}{:}\bigr]\; \tau_{11}^{n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m}^1{:}\bigr]\; \tau_{1m}^1, \,\ldots,\, \big[\const{s}_{1m}^{n_{1m}}{:}\big]\; \tau_{1m}^{n_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i$
and $\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}$. Defaults are assumed for
the selector names if they are omitted.

@@

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{gllist}(\alpha) & \,=\, \const{GLNil} \,\mid\, \const{GLCons}(\alpha,\, \ty{gllist}(\alpha)) \\[-.5\jot]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{gllist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here. In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes, and
likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion (as in the \ty{tree} example above) as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).

Because the types are inhabited (nonempty), a datatype specification is
admissible only if a ground term can be exhibited that belongs to that datatype.
This rules out ill-founded specifications such as
%
\[\keyw{datatype}\;\, \ty{fstream} \,=\, \const{FSCons}(\ty{int},\,\ty{fstream})\]
%
For codatatypes, no such check is necessary. Given the specification
%
\[\keyw{codatatype}\;\, \ty{stream} \,=\, \const{SCons}(\ty{int},\,\ty{stream})\]
%
the infinite value $\const{SCons}(0, \const{SCons}(0, \ldots))$ witnesses
nonemptiness.

@@

@@

\subsection{Correctness}

@@

@@

  * more evaluation
    * use it in a new Nitpick-like tool
  * recursive and corecursive functions
    (and rewriting)
  * extensions, perhaps straightforward, to polymorphism and nested
    (co)recursion
  * quantifiers?

>>>>

\newcommand\cpp{C\nobreak\raisebox{.05ex}{+}\nobreak\raisebox{.05ex}{+}}

@@

\newcommand\Data{\Types_{\mathrm{dt}}}
\newcommand\Codata{\Types_{\mathrm{cdt}}}
\newcommand\Nondata{\Types_{\mathrm{oth}}}

@@

Codatatypes, like datatypes, are freely generated by their constructors;
but for them, infinite constructor terms are also legitimate values
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all finite and infinite ground constructor terms, and
only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{S}(\infty) \teq \infty$ hold without any additional
work, because both sides share the same infinite expansion
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies a
codatatype value.

@@

In this paper, we introduce a unified decision procedure for ground problems
involving both datatypes and codatatypes. The procedure is described abstractly
as a calculus (Section~\ref{sec:the-calculus}) and is design to be composable
using the Nelson--Oppen method \cite{nelson-oppen-1979}. To our knowledge, this
is the first such decision procedure for codatatypes. For the datatype case, it
roughly follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.

\newcommand\cheat{\vskip.75\smallskipamount} %% TYPESETTING

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. The main differences are as follows:
%
\begin{itemize}
\item
No well-foundedness check; e.g. the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed, whereas the corresponding datatype would be rejected, because no
finite ground constructor terms exist.

\cheat
\item
The cyclicity check must be dropped. A constraint such as
$\const{x} \teq \const{C}(\const{x})$, where $\const{C}$ is a constructor,
is unsatisfiable for datatypes but satisfiable for codatatypes.
Instead: bisimilarity check, which states that two values having the same (possibly
infinite) expansion must be equal. From $\const{x} \teq \const{C}(\const{y})$ and
$\const{y} \teq \const{C}(\const{x})$, deduce $\const{x} = \const{y}$.

\cheat
\item
Normal forms are needed and must be enumeratable. For finite model finding, requires a $\mu$ notation,
e.g., $\const{SCons}(0,\, \mu s.\; \const{SCons}(1,\, \const{SCons}(2,\, s)))$
represents the lasso-shaped sequence $0, 1, 2, 1, 2, \ldots$.
\end{itemize}

Both cyclicity and bisimilarity checks cannot be replaced by finite
axiomatizations, so they really belong in a decision procedure. They are
necessary for completeness of ground problems and for soundness of found models.
%    * in particular, acyclicity and bisimilarity are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
 And the other
(co)datatype properties, such as distinctness of constructors and
exhaustiveness, can be done more efficiently by a decision procedure than by an
axiom.
%
The proofs or soundness and completeness are included in the technical report
associated with this paper.

The procedure is implemented in CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}). It implements a precise strategy for applying
inferences. It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the 1.5 release.
%
The decision procedure is evaluated on %hand-crafted examples and on
problems generated from Isabelle theories using Isabelle's Sledgehammer tool
(Section~\ref{sec:experimental-results}).

%  * useful both for proving and for model finding

%  * benchmarking is often an issue -- esp. codatatypes
%
%Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall

%  * setting: FOL
%    * core procedure is restricted to ground
%      * but theory solver cooperates
%    * still, universal conjectures fall into the ground fragment via negation
%      and skolemization (or, alternatively, herbrandization)

@@

%  * setting :
%    * universal formulas
%    * many-sorted logic
%   * mutually (co)recursive types with constructors, selectors, and
%      discriminators

%  * codatatypes: from a theoretical and implementational point of view, like
%      datatypes but:
%    * infinite values (infinitely many nested constructors)
%    * codatatypes are never empty (e.g. finite streams are rejected)

%\[
%      \keyw{codatatype}\; \,\ty{llist} \,=\, \const{LNil} \,\mid\, \const{LCons}(\ty{int},\, \ty{llist})
%\]

%  * consider a natural fragment---datatypes as supported in modern SMT solvers
%    and the SMT-LIB 2 standard, and codatatypes as their duals
%    * mutual recursion, but no polymorphism, nested recursion
%  * integrated with Nelson-Oppen

% * SMT-Lib

@@

%  * perhaps the most commonly used codatatype is that of lazy lists or sequences.
%    using a syntax similar to Standard ML, Haskell, or SMT-LIB

%  * less briefly: codatatypes and why they are useful
%    * in Agda, Coq, Matita -- recently also in Isabelle/HOL \cite{nipkow-et-al-2002}
%    * but not in SMT-LIB 2 \cite{barrett-et-al-2010}

% \paragraph{Related Work.}
% 
%     * point to Barrett et al. for SMT datatypes
%       * about their own work, they say: "our focus is on generality and
%         efficiency rather than immediacy of implementation"
%       * (deal more directly with finite sorts than Barrett et al., Section 6.1)
%     * SPASS-Pirate
%     * additional ones since then (e.g. strings?)
%     * anything about codatatypes?
%       * proof assistants like Agda, Coq, etc. have them
%       * Dafny, CoALP
%       * also a lot of theoretical research, some of which is loosely connected,
%         e.g. decision procedure for corecursive functions (Henning in Nijmegen)
% 
%     * ODDITY: Oppen 1980: single-constructor, recursive -- infinite values?

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment. In
the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
by using subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types $\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$, etc.

Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes.
%
To every argument corresponds a selector. The introduced names for the (co)data\-types, the
constructors, and the selectors must be distinct and different from already
introduced ones.%
\footnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{11}^1{:}\bigr]\; \tau_{11}^1, \,\ldots,\, \bigl[\const{s}_{11}^{n_{11}}{:}\bigr]\; \tau_{11}^{n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m}^1{:}\bigr]\; \tau_{1m}^1, \,\ldots,\, \big[\const{s}_{1m}^{n_{1m}}{:}\big]\; \tau_{1m}^{n_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i$
and $\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}$. Defaults are assumed for
the selector names if they are omitted.

@@

Here are a few examples of legal specifications of (co)datatype families:
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree}_{\,\tau} & \,=\, \const{Node}(\tau,\, \ty{forest}_{\,\tau}) \\[-\jot]
      \keyw{and}\;\, \ty{forest}_{\,\tau} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree}_{\,\tau},\, \ty{forest}_{\,\tau}) \\%[-.5\jot]
      \keyw{codatatype}\; \,\ty{llist}_{\,\tau} & \,=\, \const{LNil} \,\mid\, \const{LCons}(\tau,\, \ty{llist}_{\,\tau})
\end{aligned}
\]

Because the types are inhabited (nonempty), a datatype specification is
admissible only if a ground term can be exhibited that belongs to that datatype.
%This rules out ill-founded specifications such as
%
%\[\keyw{datatype}\;\, \ty{fstream} \,=\, \const{FSCons}(\ty{int},\,\ty{fstream})\]
%
For codatatypes, no such check is necessary because there is always a term,
finite or infinite, that witnesses nonemptiness.
%Given the specification
%
\[\keyw{codatatype}\;\, \ty{stream} \,=\, \const{SCons}(\ty{int},\,\ty{stream})\]
%
%the infinite value $\const{SCons}(0, \const{SCons}(0, \ldots))$ witnesses
%nonemptiness.

@@

\subsection{The Ground Case}

@@

\paragraph{Correctness.}

@@

\subsection{Extension to Quantified Formulas}

Blah.

@@

  * more evaluation
    * use it in a new Nitpick-like tool
  * recursive and corecursive functions
    (and rewriting)
  * extensions, perhaps straightforward, to polymorphism and nested
    (co)recursion

In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes, and
likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion (as in the \ty{tree} example above) as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).

  * quantifiers?
